---
title: "third try"
output: html_document
date: "2025-01-10"
---

---
title: "Principal Component and Clustering Analysis of Tumor Characteristics"
author: "Otajon Yuldashev"
output: html_document
date: "2025-01-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The **Breast Cancer Wisconsin (Diagnostic) Dataset** is a well-known dataset used for binary classification tasks related to breast cancer diagnosis. The dataset contains features computed from breast tissue samples collected via **Fine Needle Aspiration (FNA)**. The goal is to distinguish between **benign (non-cancerous)** and **malignant (cancerous)** tumors based on various measurements of cell nuclei.

---

### General Information
- **Dataset Name:** Breast Cancer Wisconsin (Diagnostic)
- **Source:** UCI Machine Learning Repository
- **Number of Instances (Rows):** 569 samples
- **Number of Features (Columns):** 32 total columns:
  - **1 target variable** (`diagnosis` – Benign or Malignant)
  - **30 numerical features** related to tumor characteristics.
  - **1 identifier column** (patient ID)
  

## Load Necessary Libraries
```{r,warning=FALSE}
library(tidyverse)
library(factoextra)
library(GGally)
library(corrplot)
```

## Load the Dataset
The dataset is loaded from the specified file path:
```{r}
data <- read.csv("C:\\Users\\user\\OneDrive\\Desktop\\pca project final\\data.csv")
```

## View Structure, Summary, and Missing Values
```{r}
str(data)
summary(data)
sum(is.na(data))  
```
**Observations:** The dataset contains irrelevant columns such as patient IDs (first column) and a column full of missing values (last column).

## Data Cleaning
### Step 1: Remove Irrelevant Columns
The first column (patient IDs) and the last column (containing only NA values) are removed:
```{r}
data <- data[, -c(1, ncol(data))]
```

Verify the structure:
```{r}
str(data)
head(data)
```

###  Handling Zero Values
While observing the middle and bottom sides of the dataset I came across some rows with `0` values. So in order to make our results more reliable , I'm gonna remove all rows with `0` values.So lets first check how many rows we have in total , and then remove the rows with `0` values.
Check how many rows contain zeros:


```{r}
# Step 1: Remove rows with zero values across all columns
filtered_data <- data %>% filter(if_all(everything(), ~ . != 0))

# Step 2: Select only the numerical columns after filtering
numeric_data <- filtered_data %>% select(where(is.numeric))

# Verify the structure of the cleaned numeric data
str(numeric_data)
print(paste("Number of rows after removing rows with 0:", nrow(numeric_data)))

```


## Descriptive Statistics
### Summary of Numeric Data
```{r}
summary(numeric_data)
```

## Summary of Numeric Data

The dataset contains different numerical features that describe tumor characteristics. Below is an overview of what the data tells us:

### 1. Size-Related Features:
- **`radius_mean` (Average size of the tumor’s radius):**  
  - Values range from **7.69 to 28.11**, with an average of **14.24**.  
  - The median is **13.45**, which is close to the mean, showing the distribution isn’t heavily skewed, but there are some bigger tumors.
  
- **`perimeter_mean` and `area_mean` (Perimeter and area of tumors):**  
  - **Perimeter** ranges from **48.34 to 188.50**.  
  - **Area** ranges from **170.4 to 2501.0**—this is a huge range, meaning some tumors are much larger than the others, which could indicate outliers.

---

### 2. Shape-Related Features:
- Features like **`smoothness_mean`, `compactness_mean`, `concavity_mean`, and `concave.points_mean`** describe the tumor’s shape:
  - **`smoothness_mean`** (how smooth the tumor surface is) has small values ranging from **0.05 to 0.30**, meaning there isn’t a lot of difference between tumors in this aspect.
  - **`concavity_mean`** (how concave the tumor’s border is) shows more variability, meaning some tumors have more irregular shapes.

---

### 3. Texture-Related Features:
- **`texture_mean` (Average texture of the tumor):**  
  - Values range from **9.71 to 39.28** with an average of **19.26**.  
  - The values seem fairly balanced since the mean and median are similar.

---

### 4. Complexity of Tumor Border:
- **`fractal_dimension_mean` (a measure of how complex the tumor’s border is):**  
  - Values are very small (between **0.04** and **0.30**), meaning the complexity is generally consistent across tumors.

---

### 5. Overall Spread of the Data:
- Most features have a large range (difference between the minimum and maximum values), especially for **size-related features** like `radius_mean`, `perimeter_mean`, and `area_mean`.
- This shows that tumors can differ a lot in size and shape.

---

### 6. Outliers:
- Some variables, like `area_mean`, have very large maximum values, which could mean there are **outliers**—these could be unusually large tumors.

---

### **Conclusion:**
We observe that tumors can vary a lot in size, shape, and texture. The variables related to size, like `area_mean` and `perimeter_mean`, have big ranges, while shape-related variables like `smoothness_mean` don’t vary as much. There may also be some outliers that we should keep in mind for further analysis. This makes it a good dataset for techniques like **PCA and clustering** to understand patterns in the tumor characteristics.

---



### Visualize Distributions with Boxplots
This boxplot shows the distribution of values for each variable before scaling:

```{r}
numeric_data %>%
  gather(variable, value) %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Boxplot of Variables Before Scaling")
```

#### **Description of the Boxplot:**
- **Variable Range Differences:**
  - Some variables, such as `area_mean`, `area_se`, and `area_worst`, have significantly higher values (ranging up to 4000).
  - Other variables, such as `fractal_dimension_mean`, `concavity_mean`, and `symmetry_se`, have values closer to 0 or below 1.

- **Outliers:**
  - There are noticeable outliers, particularly in `area_mean` and related variables, representing samples with extreme values for the respective feature.

- **Scale Imbalance:**
  - The wide range of scales across different variables indicates that certain features (e.g., tumor size-related metrics) will dominate the analysis unless the data is standardized.

#### **Overview for PCA:**
- PCA requires that all variables be on a similar scale to ensure fair comparison and contribution to the components.
- Variables with larger scales can heavily influence the direction of the principal components, leading to skewed results.
- After scaling, the variables will have a mean of 0 and a standard deviation of 1, equalizing their contributions to the PCA.





## Data Standardization
Standardize the data:
```{r}
scaled_data <- scale(numeric_data)
scaled_data <- as.data.frame(scaled_data)
```

## Correlation Heatmap
The heatmap shows the correlation matrix of the variables:
```{r}
corr_matrix <- cor(numeric_data)
corrplot::corrplot(corr_matrix, method = "color", tl.cex = 0.7)
```

## Correlation Heatmap Interpretation

The heatmap above represents the **correlation matrix** of the variables in the dataset. Each square in the heatmap shows the correlation between two variables, with the color indicating the strength and direction of the correlation.


1. **Clustered Correlations:**
   - Groups of variables with strong positive correlations (dark blue clusters) suggest that these variables are closely related and may contribute similarly to the overall variance.
   - For example, in the PCA context, highly correlated variables tend to have similar contributions to the principal components.

2. **Overview for PCA:**
   - Variables that show strong correlations can be combined into fewer components in PCA to reduce redundancy while still retaining most of the information.
   - If variables are uncorrelated, they may contribute unique information, and their individual influence will be more spread across different principal components.

---

### Conclusion:
The heatmap shows that many variables are highly correlated, suggesting that PCA is a suitable technique for dimensionality reduction as it captures correlated variables within fewer principal components.

---


## Principal Component Analysis (PCA)
Perform PCA:
```{r}
pca_result <- prcomp(scaled_data, center = TRUE, scale. = TRUE)
summary(pca_result)
```


## PCA Summary Table Interpretation

The table above shows the **importance of components** from the Principal Component Analysis (PCA). It provides details for each **principal component (PC)**, including the standard deviation, proportion of variance explained, and cumulative proportion of variance explained.

### Key Sections:

1. **Standard Deviation:**
   - This represents the square root of the eigenvalues associated with each principal component.
   - A higher standard deviation means that the principal component explains more variance.
   - For example, **PC1** has a standard deviation of **3.6392**, indicating that it captures much more information compared to later PCs such as **PC30** (with a standard deviation of **0.01173**).

2. **Proportion of Variance:**
   - This shows the proportion of the total variance explained by each principal component.
   - **PC1 explains 44.1%** of the total variance, **PC2 explains 19.3%**, and **PC3 explains 9.5%**.
   - Subsequent PCs explain less variance, with very small contributions from PCs beyond **PC10**.

3. **Cumulative Proportion:**
   - This represents the cumulative sum of the variance explained by the principal components.
   - Key cumulative proportions:
     - The first two components (PC1 and PC2) explain **63.4%** of the variance combined.
     - The first four components (PC1 to PC4) explain **79.4%** of the total variance.
     - By including up to **PC7**, you capture over **91%** of the variance.

### **Overview for Dimensionality Reduction:**

- The first few components (PC1 to PC4) explain most of the variance in the data and are therefore the most informative.
- Components with very low contributions (e.g., **PC20 and beyond**) add minimal information and can likely be discarded.
- A common threshold for deciding how many PCs to retain is to aim for **80-90% cumulative variance**. In this case, retaining **4 to 7 components** would capture the majority of the variance while significantly reducing the dimensionality.

---


### **Conclusion:**
For this dataset, I could retain approximately **4 to 7 components** to capture a significant amount of variance while discarding the rest to simplify the data.


---

## Scree Plot
The scree plot shows the percentage of variance explained by each component:
```{r}
  fviz_eig(pca_result, addlabels = TRUE) +
  labs(title = "Scree Plot: Variance Explained by PCs", x = "Principal Components", y = "Percentage of Variance Explained")
```

## Scree Plot Interpretation

This scree plot shows the **percentage of variance explained by each principal component (PC)** in the Principal Component Analysis (PCA).

### Key Points:

1. **X-Axis (Principal Components):**  
   The X-axis represents the different principal components (PC1, PC2, PC3, etc.). Each principal component is an independent linear combination of the original variables.

2. **Y-Axis (Percentage of Variance Explained):**  
   The Y-axis shows how much variance each principal component explains as a percentage of the total variance in the dataset.

### Interpretation:

1. **Explained Variance by PC1 and PC2:**
   - **PC1 explains 44.1%** of the total variance.
   - **PC2 explains 19.3%** of the total variance.
   - Together, the first two PCs explain **63.4%** of the total variance, which is a substantial amount.

2. **Elbow Point (Dimensionality Reduction Insight):**  
   The "elbow" point appears to be around **PC2 to PC3**. After this point, the explained variance gradually decreases and becomes smaller for the subsequent components.Suggesting that using the first few components (e.g., PC1 and PC3) may capture most of the essential information in the data, making them a good choice for dimensionality reduction.

3. **Remaining Components:**  
   The later components (e.g., PC7 onward) explain very little variance individually (close to 1-2%). These components may not add significant value to the overall understanding of the dataset.


--- 


### **Conclusion:**

- The first few principal components (e.g., PC1 to PC3) capture most of the dataset’s variability.
- Discarding the later components with minimal variance (e.g., PC7 onward) can simplify the dataset without significant information loss.


---

## Contributions of Variables to Principal Components
### Top 10 Contributing Variables to PC1, PC2, and PC3
```{r}
contribution_table <- as.data.frame(pca_result$rotation[, 1:3]) %>%
  rownames_to_column("Variable") %>%
  arrange(desc(abs(PC1))) %>%
  slice(1:10)
contribution_table
```


### **1. Top 10 Variable Contributions to PC1**

This section describes the contributions of variables to the first principal component (PC1).


```{r}
# Top 10 variables contributing to PC1
fviz_contrib(pca_result, choice = "var", axes = 1, top = 10) +
  ggtitle("Top 10 Variable Contributions to PC1")
```

#### **Description:**
- **Main Contributors:**
  - `concave.points_mean`, `concavity_mean`, and `concave.points_worst` have the highest contributions, each contributing over **6%**.
  - `compactness_mean`, `perimeter_worst`, and `concavity_worst` also contribute significantly.
- **Overview:**
  - PC1 captures variability related to **tumor shape and size**, focusing on **concavity and perimeter-related measurements**.
  - The dominance of `concave.points_mean` and related metrics suggests that PC1 reflects variations in **tumor protrusions and point distribution characteristics**.

---

### **2. Top 10 Variable Contributions to PC2**

This section describes the contributions of variables to the second principal component (PC2).

```{r}

# Top 10 variables contributing to PC2
fviz_contrib(pca_result, choice = "var", axes = 2, top = 10) +
  ggtitle("Top 10 Variable Contributions to PC2")
```

#### **Description:**
- **Main Contributors:**
  - `fractal_dimension_mean` is the highest contributor, with over **10%** contribution, followed by `fractal_dimension_se` and `fractal_dimension_worst`.
  - Other important contributors include `radius_mean`, `compactness_se`, and `area_mean`.
- **Overview:**
  - PC2 primarily captures **border complexity and irregularity**.
  - The dominance of `fractal_dimension` variables suggests that PC2 represents variations in **tumor border fractality**, emphasizing how rough or complex the tumor's boundary is.

---

### **3. Top 10 Variable Contributions to PC3**

This section describes the contributions of variables to the third principal component (PC3).


```{r}
# Top 10 variables contributing to PC3
fviz_contrib(pca_result, choice = "var", axes = 3, top = 10) +
  ggtitle("Top 10 Variable Contributions to PC3")
```

#### **Description:**
- **Main Contributors:**
  - `texture_se` contributes the most, with over **12%** of the total contribution.
  - Other significant contributors include `smoothness_se`, `symmetry_worst`, and `smoothness_worst`.
  - Additional contributors include `concave.points_se`, `radius_se`, and `compactness_worst`.
- **Overview:**
  - PC3 captures variability related to **surface texture and smoothness**.
  - The strong contributions of `texture_se` and `smoothness_se` indicate that PC3 reflects variations in **the uniformity and surface granularity** of tumor regions.
  - The presence of `symmetry_worst` suggests that this component may also capture asymmetry in extreme cases.

---

### **General Insights:**
- **PC1:** Focuses on **tumor shape and size**, emphasizing concavity and perimeter features.
- **PC2:** Represents **border complexity**, particularly variations in **fractal dimensions**.
- **PC3:** Captures **surface texture and measurement variations**, highlighting standard errors in texture and smoothness.


---


## PCA Biplot of Variables
The biplot shows the contributions of variables in the PCA space:
```{r}
  fviz_pca_var(pca_result, col.var = "darkred",labelsize=2) +
  labs(title = "Variables - PCA",
       x = paste0("Dim1 (", round(summary(pca_result)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("Dim2 (", round(summary(pca_result)$importance[2, 2] * 100, 1), "%)"))
```

### **Overview:**

1. **Axes (Dim1 and Dim2):**
   - The X-axis (Dim1) represents the first principal component (PC1), which explains **44.1%** of the variance.
   - The Y-axis (Dim2) represents the second principal component (PC2), explaining **19.3%** of the variance.
   
2. **Variable Arrows:**
   - Each arrow represents a variable in the dataset.
   - The **length of the arrow** indicates the strength of the contribution of the variable to the components.
   - Variables with longer arrows contribute more to the variance explained by the PCs.
   
3. **Direction of Arrows:**
   - Arrows pointing in similar directions indicate that the variables are positively correlated.
   - Arrows pointing in opposite directions indicate a negative correlation.
   - Variables that form a 90-degree angle (perpendicular) are uncorrelated.

4. **Relationships:**
   - Variables such as `concavity_mean`, `concave.points_mean`, and `compactness_mean` have long arrows along **PC1**, indicating that PC1 is strongly influenced by tumor shape-related features.
   - In contrast, `fractal_dimension_mean` has a strong contribution along **PC2**, suggesting that this principal component distinguishes samples based on complexity-related features.
   
5. **Circle of Correlations:**
   - The circle helps interpret the relationships between variables. Variables close to the edge of the circle have a stronger contribution, while variables closer to the origin (center) contribute less to the components.

---

### **Conclusion:**
- The biplot shows that **tumor shape metrics** dominate the variance along PC1, while **border complexity metrics** such as `fractal_dimension` contribute heavily to PC2.
- This visual reinforces the insights from the scree plot by showing which features drive the separation along the principal components.


--- 


# Prepare PCA results for clustering (first two principal components)
```{r}

pca_2d <- data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2])

# Perform K-means clustering on PCA results
set.seed(123)  # For reproducibility
kmeans_pca <- kmeans(pca_2d, centers = 2, nstart = 25)  # Choose 2 clusters (e.g., benign vs malignant)

# Add cluster labels to PCA data
pca_2d$Cluster <- as.factor(kmeans_pca$cluster)

# Visualize K-means clustering on PCA results
ggplot(pca_2d, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "K-means Clustering on PCA (2D)", x = "PC1", y = "PC2") +
  theme_minimal() +
  scale_color_manual(values = c("1" = "blue", "2" = "red"))  # Cluster 1 = blue, Cluster 2 = red

```

### Description of PCA Clustering Result

The following describes the results of **K-means clustering** applied to the **first two principal components (PC1 and PC2)** obtained from **Principal Component Analysis (PCA)**.

### Overview
- The separation between the clusters in the plot shows how the data points differ in terms of their principal components.
- The blue and red points are relatively well-separated, indicating that the K-means algorithm identified two distinct clusters based on the transformed PCA space.
- This analysis helps understand how well the original features, reduced to two dimensions, contribute to the differences in diagnosis types.


---

## Clustering the PCA results in terms of categorical variable(Diagnosis)

```{r}
# Add diagnosis labels to PCA data
pca_2d$Diagnosis <- filtered_data$diagnosis

# Visualize clusters with diagnosis labels
ggplot(pca_2d, aes(x = PC1, y = PC2, color = Diagnosis, shape = Cluster)) +
  geom_point(size = 3) +
  labs(title = "K-means Clustering on PCA (2D) with Diagnosis Comparison", x = "PC1", y = "PC2") +
  theme_minimal() +
  scale_color_manual(values = c("B" = "green", "M" = "red"))  # Diagnosis: B = green, M = red

```


### Description of PCA Clustering Result with Diagnosis Comparison

The following describes the **K-means clustering results** with an overlay of **diagnosis labels** for comparison, applied to the **first two principal components (PC1 and PC2)** obtained from **Principal Component Analysis (PCA)**.

### Overview
- The clustering results (shapes) were compared with the actual diagnosis (colors):
  - Cluster 1 (circles) mostly corresponds to benign observations.
  - Cluster 2 (triangles) corresponds to malignant observations.
- Some overlap exists, but the clustering generally aligns well with the actual diagnosis.
- This comparison highlights how well K-means clustering, applied to the PCA-reduced dimensions, captures the distinction between benign and malignant diagnoses.

---

## Computing t-SNE

### What is t-SNE?

**T-Distributed Stochastic Neighbor Embedding (t-SNE)** is a **non-linear dimensionality reduction** technique primarily used for **visualizing high-dimensional data** in a low-dimensional space (typically 2D or 3D). Developed by Laurens van der Maaten and Geoffrey Hinton, t-SNE preserves the **local structure** of data by mapping similar points in high-dimensional space to nearby points in the low-dimensional representation.

### How t-SNE Works:
1. t-SNE models the **similarity between pairs of points** in the original high-dimensional space using a conditional probability distribution.
2. In the low-dimensional space, it attempts to **recreate those similarities** by minimizing a loss function called **Kullback-Leibler (KL) divergence**.
3. t-SNE gives more weight to preserving local neighborhoods, making it particularly good for visualizing clusters.

---

### Why t-SNE Was Used in This Project

t-SNE was applied in addition to PCA for the following reasons:

1. **Capture Non-linear Relationships:**  
   PCA assumes that variance in the data can be explained using linear combinations of features. However, the tumor features may exhibit complex, non-linear patterns that PCA may not fully capture. t-SNE can capture these non-linear patterns and provide more meaningful clusters.

2. **Visualize High-Dimensional Data in 2D:**  
   t-SNE is designed for visualizing high-dimensional data by mapping it to a 2D space while preserving local neighborhoods. This makes it easier to observe and interpret clusters related to different diagnoses.

3. **Complementary to PCA:**  
   PCA provides a global summary of variance in the data, whereas t-SNE focuses on preserving local relationships between data points. The use of both methods allows for a more comprehensive understanding of the data structure.

4. **Improved Cluster Separation:**  
   t-SNE is known for creating **distinct visual clusters**, which is helpful for confirming whether benign and malignant tumors form separable groups. This helps validate the clustering results observed in PCA.

---



```{r , warning=FALSE}
# Install Rtsne if not installed
if (!requireNamespace("Rtsne", quietly = TRUE)) {
  install.packages("Rtsne")
}

# Load the necessary library
library(Rtsne)

# Perform t-SNE
tsne_result <- Rtsne(scaled_data, dims = 2, perplexity = 30, max_iter = 3000)
tsne_2d <- data.frame(tsne_result$Y)
colnames(tsne_2d) <- c("tSNE1", "tSNE2")

# Add diagnosis labels to the t-SNE result
tsne_2d$Diagnosis <- filtered_data$diagnosis

# Plot the t-SNE with diagnosis labels (color-coded)
ggplot(tsne_2d, aes(x = tSNE1, y = tSNE2, color = Diagnosis)) +
  geom_point(size = 3) +
  labs(title = "t-SNE with Diagnosis Labels", x = "tSNE1", y = "tSNE2") +
  scale_color_manual(values = c("B" = "green", "M" = "red")) +
  theme_minimal()

```


### Description of t-SNE with Diagnosis Labels

### Data Preparation
- The t-SNE algorithm was applied to the **scaled dataset** with the following parameters:
  - **Dimensions (`dims`)**: 2 (two-dimensional projection)
  - **Perplexity**: 30 (controls the balance between local and global aspects)
  - **Maximum iterations (`max_iter`)**: 3000
- The resulting t-SNE components were stored as `tSNE1` and `tSNE2`.
- Diagnosis labels (`B` for benign, `M` for malignant) were added to the t-SNE results.


### Overview
- The t-SNE plot shows a clear separation between the benign (green) and malignant (red) diagnoses.
- The clusters formed in the t-SNE space suggest that the t-SNE method effectively captured the differences in the data based on diagnosis.
- Some degree of overlap still present, indicating complexity in the structure of the dataset, for example .
- This plot demonstrates that the dimensionality reduction using t-SNE preserves the separability of diagnoses when visualized in two dimensions.

---

## Overall Conclusion

This research utilized Principal Component Analysis (PCA), K-means clustering, and t-Distributed Stochastic Neighbor Embedding (t-SNE) to analyze tumor characteristics and differentiate between benign and malignant tumors. The primary goal was to reduce the dataset’s dimensionality while retaining the key patterns and insights necessary for accurate classification.

The study demonstrated that:
- PCA effectively reduced the dataset's dimensions by capturing the majority of the variance within a few principal components.
- Clustering using the PCA-reduced data identified meaningful groups corresponding to the tumor diagnosis labels.
- t-SNE provided a more visually distinct separation between benign and malignant tumors, reinforcing the findings from PCA and clustering.

This analysis confirmed the importance of specific tumor features, such as **concavity**, **perimeter**, and **fractal dimension**, in differentiating between benign and malignant tumors. Although PCA provided a linear reduction of dimensions, t-SNE's non-linear approach was able to better capture complex patterns in the data.

---

## Conclusion and Future Directions

The analysis highlighted the effectiveness of dimensionality reduction techniques like PCA and t-SNE in understanding complex datasets. K-means clustering, when applied to PCA-reduced data, provided meaningful insights into tumor classification. However, future work could explore:
- Experimentation with different **t-SNE perplexity values** to further validate the robustness of the results.


This research provides a strong foundation for applying dimensionality reduction and clustering in biomedical datasets, showcasing the potential for identifying critical features that contribute to medical diagnoses.


## Additional Resources 
- [Serrano.Academy about PCA](https://youtu.be/g-Hb26agBFg?feature=shared)
- [Statistics Globe about PCA](https://youtu.be/mNpBrHwOCt4?feature=shared)
- [StatQuest YouTube Video on PCA](https://www.youtube.com/watch?v=FgakZw6K1QQ)
- [StatQuest YouTube Video on t-SNE](https://youtu.be/NEaUSP4YerM?feature=shared)
- [Distill Guide on t-SNE](https://distill.pub/2016/misread-tsne/)
https://youtu.be/NEaUSP4YerM?feature=shared